---
title: "**MATH 3190 Homework 1**"
subtitle: |
  Due January 29, 2026\
  Focus: Notes 1-3\
  Total Points: 90 + 10 (Self-Critique Comment and Uploading to GitHub) = 100
format: html
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---

Note that your homework should be completed in Quarto. You can just add your answers and code to this document in the appropriate part. The file should be rendered as an html document. You will "turn in" this homework by uploading the files, the .qmd and output file, to your GitHub `Math_3190_Assignment` repository in the `Homework/Homework_1` directory. Make sure all supporting files (if there are any) are in this directory as well. You should use `git` and `Unix` commands in a terminal for this. In addition, you will submit just your html or pdf file on Canvas to unlock the solutions. 

Don't forget to also put a [self-critique comment]{.underline} on Canvas indicating what you did correctly and incorrectly and include any questions you may have about the assignment after submitting the assignment and thoroughly looking through the solutions.


# Problem 1 (18 points)
```{r, echo=TRUE, eval=TRUE}
#| label: gghist
library(ggplot2)

gghist <- function(x){
  if (!is.numeric(x)) {
    stop("Please enter a numeric vector")
  }
  df <- data.frame(x = x)
  
  ggplot(df, aes(x = x)) +
    geom_histogram(
      binwidth =0.7,
      color ="black",
      fill ="lightblue") +
    labs(y ="Count", x ="Input", title ="Histogram")
}
```

```{r, echo=TRUE, eval=TRUE}
#| label: ggbox
library(ggplot2)

ggbox <- function(x, group =NULL, orientation = c("horizontal", "vertical")){
  
  if (!is.numeric(x)){
    stop("Please enter a numeric vector")
  }
  
  if (is.null(group)){
    y <- data.frame(x = x)
    
    g <- ggplot(y, aes(y = x)) +
      geom_boxplot(fill ="lightgreen") +
      labs(x ="", y ="")
  } else {
    y <- data.frame(x = x, group = group)
    
    g <- ggplot(y, aes(x =group, y =x)) +
      geom_boxplot(fill ="pink") +
      labs(x ="", y="")
  }
  
  if (orientation == "horizontal"){
    g <- g + coord_flip()
  }
  
  return(g)
}
```

# Problem 2 (57 points)

### Part a (6 points)
```{r}
#| label: read_bb_scores

library(tidyverse)

cbbga <- tibble(
  read.fwf("http://kenpom.com/cbbga26.txt",
           widths =c(10, 24, 3, 24, 3, 20),
           strip.white =TRUE)
  )
```

### Part b (2 points)
```{r}
#| label: Problem 2b

cbbga2 <- cbbga |> 
rename(
  Date = V1, 
  Visiting_Team = V2,
  Visiting_Score = V3,
  Home_Team = V4,
  Home_Score = V5,
  Random = V6
)
head(cbbga2)
```

### Part c (2 points)
```{r}
#| label: Problem 2c

cbbga3 <- cbbga2 |>
  mutate(
    Score_Difference = Home_Score - Visiting_Score
  )
head(cbbga3)
```
### Part d (2 points)

```{r}
#| label: Problem 2d

cbbga4 <- cbbga3 |>
  arrange(Home_Team)
head(cbbga4)
```

### Part e (2 points)

```{r}
#| label: Problem 2e

cbbga5 <- cbbga4 |>
  select(!'Random')
head(cbbga5)
```
### Part f (3 points)
Put parts a-e all together in one piping expression (with 5 pipes) and save this as a new object in **R**.
```{r}
#| label: Problem 2f

fancy_scores <- 
  tibble(
    read.fwf(
      "http://kenpom.com/cbbga26.txt",
      widths =c(10, 24, 3, 24, 3, 20),
      strip.white =TRUE)
  ) |>
  rename(
    Date = V1, 
    Visiting_Team = V2,
    Visiting_Score = V3,
    Home_Team = V4,
    Home_Score = V5,
    Random = V6
  ) |>
  mutate(
    Score_Difference = Home_Score - Visiting_Score
  ) |> arrange(Home_Team) |> select(!'Random')
head(fancy_scores)
```

### Part g (3 points)
Use `filter()` to reduce the data down to only games played in 2025 (you could use the `lubridate` package for this, since it specializes in dealing with dates, but some base **R** packages will also work). Save this in a new tibble. **We will use this tibble with only the 2025 years from here on out**.
```{r}
#| label: Problem 2g

fancy_scores$Date <- as.Date(fancy_scores$Date, format = "%m/%d/%Y")

games_2025 <- fancy_scores %>% 
  filter(format(Date, "%Y") == "2025")
head(games_2025)
```

### Part h (4 points)
Write a function that will filter the tibble to only games played by a given team. Demonstrate your function by displaying games played by SUU in 2025.
```{r}
#| label: Problem 2h

team <- function(x){
  games_2025 %>% filter(Visiting_Team == x | Home_Team == x)
}
team("Southern Utah")
```

### Part i (7 points)

```{r}
#| label: Problem 2i

suu_2025 <- team("Southern Utah") %>% 
  mutate(
    result = case_when(
      Home_Team == "Southern Utah" & Home_Score > Visiting_Score ~ "Win",
      Home_Team == "Southern Utah" & Home_Score < Visiting_Score ~ "Loss",
      Visiting_Team == "Southern Utah" & Home_Score < Visiting_Score ~ "Win",
      Visiting_Team == "Southern Utah" & Home_Score > Visiting_Score ~ "Loss"
    )
  ) %>% 
  summarize(
    Wins = sum(result == "Win"),
    Losses = sum(result == "Loss"),
    Games = n(),
    win_percentage = Wins / Games
  )

suu_2025

```

### Part j (7 points)
Generalize this by writing a function that will do this for a given team, and create a tibble with this information for all teams. Arrange this tibble first by winning percentage (descending) and by team (alphabetically) if there are any ties when it comes to winning percentage. A `for` loop and the `add_row()` function may be useful here.
```{r}
#| label: Problem 2j

team_win_record <- function(x){
  games_2025 %>% 
    filter(Visiting_Team == x | Home_Team == x) %>% 
    mutate(
      win = case_when(
        Home_Team == x & Home_Score > Visiting_Score ~ 1,
        Visiting_Team == x & Visiting_Score > Home_Score ~ 1,
        TRUE ~ 0
      )
    ) %>% 
    summarise(
      Team = x,
      Wins = sum(win),
      Losses = n() - Wins,
      Games = n(),
      Win_Percentage = Wins/Games
    )
}

all_team <- sort(unique(c(games_2025$Visiting_Team, games_2025$Home_Team)))

records <- tibble(
  Team = character(),
  Wins = integer(),
  Losses = integer(),
  Games = integer(),
  Win_Percentage = numeric()
)
for (t in all_team) {
  team_stats <- team_win_record(t)
  records <- add_row(records, !!!team_stats)
}
records <- records %>% 
  arrange(desc(Win_Percentage), Team)

head(records)
```

### Part k (5 points)
With only the 2025 games, use the `gghist()` you wrote in Problem 1 on the score differences variable you created in part c. Add appropriate labels and a title to your graph. Also, use the `#| fig-width:`, `#| fig-height:`, `#| out-width:`, and `#| fig-align:` code chunk options to center the figure and then to adjust your figure to a size you like. Play around with different sizes to see what those options adjust.
```{r}
#| label: Problem 2k
#| fig-width: 7
#| fig-height: 5
#| out-width: "70%"
#| fig-align: "center"

gghist(games_2025$Score_Difference) +
  labs(
  x = "Score Difference (Home Score - Visiting Score)",
  y = "Number of Games",
  title = "Score Differences from 2025 College Basketball Games"
  )
```

### Part l (9 points)
Filter the tibble with all 2025 games to only include Utah Tech or Southern Utah, then crate a tibble or data frame where each row is a game played by either team that has one column indicating the school, Utah Tech or Southern Utah, and a second column indicating the number of points that team played, regardless of whether the team was home or away. Then use your `ggbox()` function to create a side-by-side boxplot of number of points scored in all 2025 games for each team.

Note, this is a little challenging. The most straightforward way to do this is to use the `pivot_longer()` function with `cols = ` the names of the two columns that have the away and the home teams (the names of these columns should not be in quotes, but they should be in a vector `c()`). Try to do this yourself first, but using AI for some help is fine. 

### Part m (5 points)
Create one more appropriate graphs for the 2025 basketball data. This graph could be anything you'd like as long as you use `ggplot2`, it shows something meaningful, and it is not the same as the graphs you made in parts k or l. 


# Problem 3 (15 points)
Repeat parts b-e and g of Problem 2 using Python in Quarto. First, pass the original object that you read in from the website to Python without any changes to it (you do not need to read the file from the web in Python, but you can if you'd like) and then use `pandas` (or `polars` if you'd like) to rename the columns as indicated in part b, add the columns specified in part c, arrange the data as in part d, drop the "garbage" column as in part e, and filter it down as in part g. The `pandas` functions `rename`, `assign` (instead of `mutate`), `drop` (instead of `select`) and `str.contains` (used to select the right rows) will be useful here. Be sure to follow the guide in Notes 3 to properly install Python, install the `pandas` library and to load it in **R**.
